---
title: "Analysis of ENCODE TFBS"
author: "Jay Hesselberth"
date: "December 31, 2015"
output: html_document
---

```{r read_data}
library(dplyr)
library(readr)
library(tidyr)
library(stringr)
library(ggplot2)
library(broom)

url <- 'http://hgdownload.cse.ucsc.edu/goldenPath/hg19/encodeDCC/wgEncodeRegTfbsClustered/wgEncodeRegTfbsClusteredWithCellsV3.bed.gz'

col_names <- c('chrom', 'start', 'end', 'tf.name', 'cell.id', 'cell.names')
tfbs_data <- read_tsv(url, col_names = col_names) %>% tbl_df

# filter on tier 1 cell lines
tier1 <- c('K562', 'H1-hESC', 'GM12878')
tfbs_tier1_data <- tfbs_data %>%
  select(-cell.id) %>%
  mutate(cell.names = str_split(cell.names, ',')) %>%
  unnest(cell.names) %>% 
  rename(cell.name = cell.names) %>%
  filter(cell.name %in% tier1)

peak_midpoints_raw <- tfbs_tier1_data %>%
  mutate(peak.midpoint = as.integer(start + round((end - start) / 2))) %>%
  select(-start, -end)

# select groups with more than 100 observations
peak_midpoints <- peak_midpoints_raw %>%
  group_by(chrom, tf.name, cell.name) %>%
  summarize(count = n()) %>%
  filter(count > 100) %>%
  inner_join(peak_midpoints_raw) %>%
  select(-count) %>%
  ungroup()
```

```{r tfbs_summary}
```

## Are TF binding sites randomly distributed on chromosomes? (Lag plots)

```{r tfbs_lag}
# http://www.r-bloggers.com/predictability-of-stock-returns-using-acf/
peak_lags <- peak_midpoints %>%
  group_by(chrom, tf.name, cell.name) %>%
  arrange(peak.midpoint) %>%
  mutate(midpoint.lag = peak.midpoint - lag(peak.midpoint)) %>%
  select(-peak.midpoint) %>%
  na.omit()

peak_lags  
```

Let's look a little closer at the structure of the peak_lags.

```{r plot_peak_lags}
peak_lags %>%
  ggplot(aes(midpoint.lag, color = cell.name)) +
  geom_histogram() +
  facet_wrap(~tf.name) + 
  scale_y_log10() +
  xlim(c(0,100000))
```

## Do any factors exhibit non-randomness?

We can use the `randtests::runs.test()` to assess whether any of the midpoint distributions have significant levels of non-randomness.

```{r runs_test}
library(randtests)

# this takes ~2 minutes to run
runs_tests <- peak_lags %>%
  group_by(chrom, tf.name, cell.name) %>%
  do(tidy(runs.test(.$midpoint.lag)))

runs_tests <- peak_lags %>%
  group_by(chrom, tf.name, cell.name) %>%
  filter(row_number() <= 500) %>%
  do(tidy(shapiro.test(.$midpoint.lag)))
```

Now let's check the distribution of p-values in the runs tests.

```{r runs_pvalues}
runs_tests %>%
  ggplot(aes(p.value)) +
  geom_histogram(binwidth = .05)
```

# Now let's do some correction of those p-values, converting them to q-values.

```{r runs_qvalues}
library(qvalue)

runs_tests <- runs_tests %>%
  ungroup() %>%
  mutate(q.value = qvalue(p.value)$qvalues)

runs_tests %>%
  group_by(cell.name, chrom) %>%
  summarize(signif = sum(q.value < .01)) %>%
  arrange(-signif)
```

```{r lag_acf}
acf_conf <- function(x, conf.level = 0.95, ...) {
  conf.int <- qnorm((1 + conf.level) / 2) / sqrt(x$n.used)
  conf.int
}

lag_acfs <- peak_lags %>%
  group_by(chrom, tf.name, cell.name) %>%
  do(acf_result = acf(.$midpoint.lag, plot = FALSE))

tidy_lag_acfs <- lag_acfs %>% tidy(acf_result)

acf_confs <- lag_acfs %>%
  mutate(conf.level.95 = acf_conf(acf_result),
         conf.level.99 = acf_conf(acf_result, conf.level = 0.99)) %>%
  ungroup()
```

## Are there combinations that are highly significant at all acfs?

```{r}
# set a threshold for the number that can be false
# XXX could also be a proportion
min.false <- 2
min.obs <- 10

tidy_lag_acfs %>% left_join(acf_confs) %>%
  mutate(signif = (acf > 0 & acf > conf.level.95) | (acf < 0 & acf < -conf.level.95)) %>%
  select(chrom, tf.name, cell.name, signif) %>%
  summarize(num.obs = n(),
            num.true = length(which(signif)),
            num.false = num.obs - num.true) %>%
  filter(num.obs > min.obs & num.false <= min.false) %>%
  ungroup() %>%
  select(chrom, tf.name, cell.name) %>%
  unique()

acfs_high_signif <- tidy_lag_acfs %>% left_join(acf_confs) %>%
  filter(all((acf > 0 & acf > conf.level.95) | (acf < 0 & acf < -conf.level.95))) %>%
  ungroup() %>%
  select(chrom, tf.name, cell.name) %>%
  unique()

signif_data <- acfs_high_signif %>% left_join(tidy_lag_acfs) %>% left_join(acf_confs)

ggplot(signif_data, aes(x = lag, y = acf)) + geom_bar(stat='identity', width = 0.3) +
  facet_wrap(~tf.name + chrom + cell.name) + theme_bw() +
  geom_hline(aes(yintercept = conf.level.95), data = signif_data, linetype = 'dashed', color = 'blue') + 
  geom_hline(aes(yintercept = conf.level.99), data = signif_data, linetype = 'dashed', color = 'red') 
```

```{r}

lag_plot <- function(data, facets, conf_levels = TRUE, conf_data, ...) {
  
  p <- ggplot(aes(x = lag, y = acf)) + 
    facet_wrap(facets) + 
    geom_bar(stat='identity', width = 0.3) +
    theme_bw()
  
  if (conf.levels) {
    if ("conf.level.95" %in% colnames(conf_data)) {
      p <- p + geom_hline(aes(yintercept = conf.level.95),
                       linetype = 'dashed',
                       color = 'blue', data = conf_data) +
      p <- p + geom_hline(aes(yintercept = -conf.level.95),
                          linetype = 'dashed',
                          color = 'blue', data = conf_data)
    }
    if ("conf.level.99" %in% colnames(conf_data)) {
      p <- p + geom_hline(aes(yintercept = conf.level.99),
                       linetype = 'dashed',
                       color = 'red', data = conf_data) +
      p <- p + geom_hline(aes(yintercept = -conf.level.99),
                          linetype = 'dashed',
                          color = 'red', data = conf_data)
    }
  }
  return(p)
} 
```

```{r tfbs_modeling}
```