---
title: "tidying data sets"
author: "Jay Hesselberth"
date: "December 27, 2015"
output: slidy_presentation
---

## Tidying data sets

Raw data must often be restructured to enable downstream analysis. A useful format to start with is "tidy" data.

Tidy data has a few key principles:

  - Each variable forms a column
  - Each observation forms a row
  - Each type of observational unit forms a table

## Common problems with messy data

  - Column headers are values, not variable names.
  - Multiple variables are stored in one column.
  - Variables are stored in both rows and columns.

```{r anscombe, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyr)
library(dplyr)
library(ggplot2)

anscombe_tidy_hide <- anscombe %>% tbl_df %>%
  mutate(obs = seq_len(n())) %>%
  gather(key, value, -obs) %>%
  separate(key, c('var','set'), 1) %>%
  spread(var, value) %>%
  arrange(set)
```
## The `anscombe` data sets

The `anscombe` data sets illustrate a key point of data analysis: **visualize your data**.

Why? Standard statistics do not differentiate among the data sets:

```{r anscombe_stats, echo = FALSE, message = FALSE}
anscombe_tidy_hide %>%
  group_by(set) %>%
  summarize(len.data = n(),
            mean.x = round(mean(x), digits = 2),
            mean.y = round(mean(y), digits = 2),
            variance.x = round(var(x), digits = 2),
            variance.y = round(var(y), digits = 2),
            set.correlation = round(cor(x,y), digits = 2))
```

## ... but plotting them tells another story ...

```{r anscombe_plots, echo=FALSE, warning=FALSE}
anscombe_tidy_hide %>%
  ggplot(aes(x,y)) +
    geom_point() +
    facet_wrap(~set) +
    geom_smooth(method='lm', se=FALSE) +
    theme_bw()
```

## Tidying the `anscombe` data set

The provided `anscombe` data set is *messy* because the set number and x/y variables are combined in column names.

```{r echo=FALSE}
anscombe %>% tbl_df
```

## Tidying the `anscombe` data set

We need to reformat the table into this form:

```{r echo=FALSE}
anscombe_tidy_hide %>% tbl_df
```

## Adding surrogate observation counts

We need to make sure that individual observations in the tidied data set are identiable by a unique value. I call this the `observation id` or `obs`.

```{r anscombe_mutate}
anscombe_tidy <- anscombe %>% tbl_df %>%
  mutate(obs = seq_len(n()))

anscombe_tidy
```

## Using `tidyr::gather()`

```{r anscombe_gather}
anscombe_tidy <- anscombe_tidy %>% 
  tidyr::gather(key, value, -obs)

anscombe_tidy
```

Think about the transformation that just happened.

## Using `tidyr::separate()`

```{r anscombe_separate}
anscombe_tidy <- anscombe_tidy %>%
  tidyr::separate(key, into = c('var', 'set'), sep = 1)

anscombe_tidy
```

Think about the transformation that just happened.

## Using `tidyr::spread()`

We're almost there. Now we need to `spread()` the `var` and `value` columns.

```{r anscombe_spread, warning = FALSE}
anscombe_tidy <- anscombe_tidy %>%
  tidyr::spread(var, value)

anscombe_tidy
```

## Other `tidyr` verbs

Sometimes you have incomplete sets of observations in a data set. The `complete()` verb addresses this issue by filling in missing values.

```{r missing}
df <- dplyr::tibble(
  ~group, ~item_id, ~item_name, ~value1, ~value2,
  1,      1,        'a',          1,       4,
  2,      2,        'b',          2,       5,
  1,      2,        'b',          3,       6)
```

## `tidyr::complete()` the full set

```{r missing_cont, eval=FALSE}
#. 1. expand *all* cominbations
df %>% tidyr::complete(group, item_id, item_name)
```

## `tidyr::complete()` sets of variables in the given data

```{r missing_cont_2, eval=FALSE}
#. 2. only expand sets that are observed in the given data
df %>% tidyr::complete(group, c(item_id, item_name))
```

## Tidying practice

  - What happens if you retidy the `anscombe` set, but you leave out the `obs` column? 
  
  - Practice tidying other R builtin data sets. Use `library(help = "datasets")` to see all of the builtin datasets. Try the `ability.cov` dataset, and try combining the `beaver1` and `beaver2` data sets.



