---
title: "Data manipulation"
author: "Jay Hesselberth"
date: "March 6, 2015"
output: html_document
---

```{r, echo=FALSE}
library(dplyr)
library(ggplot2)
library(knitr)
```
# Data sets
There are a huge number of datasest you can play around with.

```
devtools::install_github("hadley/babynames")
```

```{r}
library(babynames)
babynames <- tbl_df(babynames)
babynames %>% head(10) %>% kable()
```

# Useful packages

## stringr

Here's how to get first letters ...

```{r}
library(stringr)
babynames %>%
  mutate(first.letter = str_sub(name, start = 1, end = 1)) %>%
  head() %>% kable()
```

Now make a plot of first letter usage ...

```{r}
babynames %>%
  mutate(first.letter = str_sub(name, start = 1, end = 1)) %>%
  group_by(sex, year, first.letter) %>%
  summarize(count = sum(n)) %>%
  ggplot(aes(x = year, y = log10(count), color = sex)) +
  geom_line() +
  facet_wrap(~ first.letter)
```

### Questions

- What is the spike in male `Q` names around 1915?

## tidyr

We'll clean up an ugly table to convert it to `tidy` format.

```{r}
library(tidyr)

letters.spread <-
  babynames %>%
    mutate(first.letter = str_sub(name, start = 1, end = 1)) %>%
    select(sex, year, first.letter, n) %>%
    group_by(year, first.letter) %>%
    summarize(count = sum(n)) %>% 
    spread(first.letter, count, fill = 0)

letters.tidy <- 
  letters.spread %>%
    gather(sex, year)
```

### Questions
- What are the most and least correlated first letters through the years? Use `cor()` and the `corrplot` package.

```{r}
library(corrplot)
M <- cor(letters.spread)

corrplot(M, method = 'circle')
```

## dplyr 

Get the top two male and female names for each year ... use `slice()`

```{r}
babynames %>%
  select(year, name, sex, n) %>%
  group_by(year, sex) %>%
  arrange(desc(n)) %>%
  slice(1:2)
```

Figure out when names changed the most ...

```{r}
# not quite working the way i intended ...
babynames %>%
  select(year, name, n) %>%
  arrange(name, n) %>%
  mutate(name_delta = name != lead(name))
```

We'll go over two-table verbs in dplyr, which allow you to join tables based on common values.